Avoiding tipping points in fisheries management through Gaussian Process Dynamic Programming

### 2014-11-20 manuscripts/manuscript.tex
as submitted

We develop a novel approach to address these concerns in the context of
fisheries; although the challenges and methods are germane to other
problems of conservation or natural resource exploitation. Economic
value and ecological concern have made marine fisheries the crucible for
much of the founding work on management under uncertainty (Gordon 1954,
Clark 1976, 2009, May et al. 1979, Reed 1979, Ludwig and Walters 1982)

@@

Tipping points in ecological dynamics (Scheffer et al. 2001, Polasky et
al. 2011) highlight this problem because precise models are not
available and data are limited such as around high stock levels or an
otherwise desirable state. With perfect information , one would know
just how far a system could be pushed before crossing the tipping point,
and management would be simple. But we face imperfect models and limited
data and, with tipping points,\\even small errors can have very large
consequences, as we shall illustrate later. Because intervention is
often (but not always, see Hughes et al. (2013)) too late after a
tipping point has been crossed, management is often concerned with
avoiding tipping points before any data about them is available or the
transition that would more clearly reveal a tipping point occurs (e.g.
Bestelmeyer et al. 2012).

@@

Dynamic optimization requires characterizing the dynamics of a state
variable (or variables), a control action, and a value function. For
simplicity, we consider only a single state variable, which allows us to
focus on one-dimensional models of population dynamics for the
comparison of parametric models and the GP. This is a best-case scenario
for the parametric models because we simulate underlying dynamics from
one of the three parametric models , whereas in the natural world we
never know the ``true'' mode . In addition, by choosing one-dimensional
models with just a few parameters, we limit the chance that poor
performance will be due to inability to estimate parameters accurately,
something that becomes a more severe problem for higher-dimensional
parametric models. Finally, the parametric models we consider are those
most commonly used in modeling stock-recruitment dynamics or to model
sudden transitions between alternative stable states.

@@

We let $X(t)$ denote the size (numbers or biomass) of the focal
population at time $t$ and assume that in the absence of take its
dynamics are. .

\begin{equation}
X(t+1) = Z(t) f(X(t), p) \label{eq1} 
\end{equation}

Where Z(t) is log-normally distributed process stochasticity (Reed 1979)
and $p$ is a vector of parameters to be estimated from the data. We
describe the three choices for $f(X(t),p)$ in the next section.\\The
control action is a harvest or take, $h(t)$, measured in the same units
as $X$, at time $t$. Thus, in the presence of take, the population size
on the right hand side of Eqn 1 is replaced by $S(t) = X(t) - h(t)$.

To construct the value function, we consider a return when $X(t) = x(t)$
and harvest $h(t) = h$ denoted as the reward, $R(x(t), h)$. For example,
if the return is the harvest at time $t$, then
$R(x(t), h(t)) = \min(x(t), h(t))$. We assume that future harvests are
discounted relative to current ones at a constant rate of discount
$\delta$ and ask for the harvest policy that maximizes total discounted
harvest between the current time $t$ and a final time $T$. That is, we
seek to maximize over choices of harvest
$E_{X(t+1)} [ \sum_{t = 0}^{T}  R(X(t), h(t), t) \delta^t]$, where the
state dynamics are given by Eqn 1 and $E$ denotes the expectation over
the process stochasticity in the future population state.

@@

We consider a period of 40 in which data are obtained to estimate the
parameters or the GP. This is long enough that the estimates do not
depend on the particular realization, and longer times are not likely to
provide substantial improvement.\\We fit each of the models is fit to
the same data (Figure 1).

@@

The core difference for our purpose between the GP and the parametric
models is that the GP model is defined explicitly by the observed data.
That is, it cannot be specified by the value of parameters alone.

@@

A GP is a stochastic process for which any realization of for a finite
sample of points $n$ is a multivariate normal distribution of dimension
$n$. Thus, to characterize the GP we need a mean function and a
covariance function. We proceed as follows.

As before, we assume that the data $X(t)$ are observed with process
stochasticity around a mean function \$ g(X(t))\$

@@

where $\varepsilon$ are IID normal random variables with zero-mean and
variance $\sigma_n^2$. Note that we have chosen to assume additive
stochasticity. While we could consider log-normal stochasticity as in
the parametric models, we make this choice to emphasize that the
Gaussian process approach need not have structurally correct
stochasticity to be effective.

Now consider the situation where we have a set of previous observations
of the \emph{timeseries}, $X_{\textrm{obs}}(t)$, and wish to
\emph{predict} future values $X_{\textrm{p}}(t)$. We consider the
observations as pairs of points
$X_{\textrm{obs}}(t), X_{\textrm{obs}}(t+1)$ for $t=1, 2, …$ up through
final observation time $T$. As described above, in the GP, a finite
sample of these predictions at $n$ points into the future,
$X_{\textrm p}(\tau)$ for $\tau \in [T+1, T+2, ..., T+n]$, is a
multivariate normal ($\mathcal{N}$):

\begin{equation}
g(X_p|X_{\textrm{obs}}(t), X_{\textrm{obs}}(t+1)) \sim \mathcal{N}(E,C)
\end{equation}

with mean and covariance matrix given by (see Eqn 2.22 of Rasmussen and
Williams (2006) for a more detailed derivation)

\begin{equation}
E = K(X_{\textrm p}(\tau), X_{\textrm{obs}}(t)) \left(K(X_{\textrm{obs}}(t),X_{\textrm{obs}}(t)) - \sigma \mathbb{I} \right)^{-1} X_{\textrm{obs}}(t+1)
\end{equation}

\begin{equation}
C = K(X_{\textrm p}(\tau), X_{\textrm p}(\tau)) - K(X_{\textrm p}(\tau), X_{\textrm{obs}}(t)) \left(K(X_{\textrm{obs}}(t),X_{\textrm{obs}}(t)) - \sigma \mathbb{I} \right)^{-1} K(X_{\textrm{obs}}(t), X_{\textrm p}(\tau))
\end{equation}

Where $K$, called the covariance kernel, measures how much influence one
observation has on another. In this paper, for $K$ use a matrix whose
$(i,j)$th element is given by:

\begin{equation}
K_{i,j}(x_i, y_j) = \exp\left( \frac{ -(x_i - y_j)^2}{2 \ell^2} \right)
\end{equation}

so that $\ell$ gives the characteristic length-scale over which
correlation between two observations decays.

We use inverse Gamma priors on both the length-scale $\ell$ and
$\sigma$, thus for example

@@

We derive the harvest policy from the estimated GP by inserting it into
a SDP algorithm. Given the GP posteriors, we construct the transition
matrix representing the probability of going to each state $X(t+1)$
given any current state $x(t)$ and any harvest $h(t)$ (See the function
\texttt{gp\_transition\_matrix()} in the provided R package). Given this
transition matrix, we use the same value iteration algorithm as in the
parametric case to determine the optimal policy.\\In doing so, the
uncertainty in the future state under the GP, $X(t+1)$, includes both
process uncertainty (based on the estimation of $\sigma$) and structural
uncertainty of the posterior collection of curves.\\Unlike the case of a
parametric model, which is a single curve given the parameters, the GP
remains a distribution of curves.

@@

To ensure our results are robust to the choice of parameters, we
consider 96 different scenarios, described in detail below. To help
better understand the process, we first describe in detail the results
of a single scenario.

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_1-1.pdf}
\caption{Points show the training data of stock-size over time. Curves
show the posterior step-ahead predictions based on each of the estimated
models. Observe that all models are fitting the data reasonably well. In
the GP panel, black lines are means of the GP for random samples from
the posterior distributions, while grey bands around them reflect the
standard deviations of the GP distribution. (Recall that while given a
particular set of parameters, a parametric model gives just one curve,
the GP is always a distribution, whose mean we plot in black.)}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_2-1.pdf}
\caption{The inferred Gaussian process compared to the true process and
maximum-likelihood estimated process. We show the expected value for the
function $f$ under each model. Two standard deviations from the
estimated Gaussian process covariance with (light grey) and without
(darker grey) measurement error are also shown. The training data are
also shown as black points. The GP is conditioned on (0,0), shown as a
pseudo-data point, as explained in the text}
\end{figure}

We show the mean inferred population dynamics of each model relative to
the true model used to generate the data in Figure 2, predicting the
relationship between observed population size (x-axis) to the population
size after recruitment the following year.\\In addition to the raw data,
the GP is conditioned on going through the point 0,0 without error. This
conditioning on (0,0) is equivalent to making the assumption that the
population is closed, so that once it hits 0 it stays at 0, despite the
lack of any data in the observed sequence to justify this. This
assumption illustrates how the GP can capture common-sense biology
without having to assume more explicit details about the dynamics at low
population numbers that have never been observed. If the population were
not closed, one could repeat the entire analysis without this
assumption. Unlike parametric models, the GP corresponds to a
distribution of curves - the gray band with mean shown in black.
Uncertainty in the parameters of the GP (not shown) further widens the
band of possible population sizes. In Figure S1 (see supplement), we
show the performance of the models outside the observed training data.

@@

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_4-1.pdf}
\caption{In the management context, GPDP outperforms approaches based on
parametric models. We show 100 replicate simulations of the stock
dynamics (Eqn 1) under the policies derived from each of the estimated
models, as well as the policy based on the exact underlying model.}
\end{figure}

@@

Our results illustrate that model-choice approaches can be absolutely
misleading -- by providing support to models that cannot capture tipping
point dynamics because they have fewer parameters and the data are far
from the tipping point. That is, when the data come from around the
stable steady state, all the parametric models are approximately linear
and approximately identical. Thus, it is intuitive that all model
selection methods choose the simplest model. In a complex world, the
result is suboptimal. But in a world that might contain tipping points,
the result is disastrous.

Many managers both in fisheries and beyond face a similar situation:
they have not observed the population dynamics at all possible
densities. A lack of comprehensive data at all population sizes,
combined with the inability to formulate accurate population models for
low population sizes in the absence of data, makes this situation the
rule more than the exception. Relying on parametric models and model
choice processes that favor simplicity ignores this basic reality. For a
long time, Carl Walters (e.g. Walters and Hilborn 1978) has argued that
if we began by fishing any newly exploited population down to very low
levels and then let it recover, we would be much better at estimating
population dynamics and thus predicting the optimal harvest levels.
While certainly true, this presents a rather risky policy in the face of
potential tipping points. The GPDP offers an alternative way to
acknowledge this uncertainty.

@@

Outside of the observed range of the data, the GP reverts to the prior,
and consequently the choice of the prior can also play a significant
role in determining the optimal policy inferred by the SDP. In the
examples shown here we have selected a prior that is both relatively
uninformative (due to the broad priors placed on its parameters $\ell$
and $\sigma$ and simple (the choice of our covariance function, Eqns 12
and 13 ). In practice, these should be chosen to confer particular
biological properties. In principle, this may allow a manager to improve
the performance of the GPDP by adding detail as is justified. For
instance, it would be possible to use a linear or a Ricker-shaped mean
in the prior without making the much stronger assumption that the Ricker
is the structurally correct model (Sugeno and Munch 2013a). One fruitful
avenue of future research is identifying criteria to ensure the prior
and the value function are chosen appropriately for the problem at hand.

@@

Bestelmeyer, B. T., M. C. Duniway, D. K. James, L. M. Burkett, and K. M.
Havstad. 2012. A test of critical thresholds and their indicators in a
desertification-prone ecosystem: more resilience than we thought.
Ecology Letters:n/a--n/a.

>>>>

We develop a novel approach to address these concerns in the context of
fisheries; although the challenges and methods are germane to other
problems of conservation or natural resource exploitation. Economic
value and ecological concern have made marine fisheries the crucible for
much of the founding work on management under uncertainty (Gordon 1954,
Clark 1976, 2009, May et al. 1979, Reed 1979, Ludwig and Walters 1982).

@@

Tipping points in ecological dynamics (Scheffer et al. 2001, Polasky et
al. 2011) highlight this problem because precise models are not
available and data are limited such as around high stock levels or an
otherwise desirable state. With perfect information , one would know
just how far a system could be pushed before crossing the tipping point,
and management would be simple. But we face imperfect models and limited
data and, with tipping points,\\even small errors can have very large
consequences, as we shall illustrate later. Because intervention may be
too late once a tipping point has been crossed (but see Hughes et al.
(2013)), management is often concerned with avoiding tipping points
before any data about them are available.

@@

Dynamic optimization requires characterizing the dynamics of a state
variable (or variables), a control action, and a value function. For
simplicity, we consider only a single state variable, which allows us to
focus on one-dimensional models of population dynamics for the
comparison of parametric models and the GP. This is a best-case scenario
for the parametric models because we simulate underlying dynamics from
one of the three parametric models , whereas in the natural world we
never know the ``true'' model. In addition, by choosing one-dimensional
models with just a few parameters, we limit the chance that poor
performance will be due to inability to estimate parameters accurately,
something that becomes a more severe problem for higher-dimensional
parametric models. Finally, the parametric models we consider are
commonly used in modeling stock-recruitment dynamics or to model sudden
transitions between alternative stable states.

@@

We let $X(t)$ denote the size (numbers or biomass) of the focal
population at time $t$ and assume that in the absence of take its
dynamics are:

\begin{equation}
X(t+1) = Z(t) f(X(t), \mathbf{p}) \label{eq1} 
\end{equation}

Where Z(t) is log-normally distributed process stochasticity (Reed 1979)
and $\mathbf{p}$ is a vector of parameters to be estimated from the
data. We describe the three choices for $f(X(t),\mathbf{p})$ in the next
section.\\The control action is a harvest or take, $h(t)$, measured in
the same units as $X$, at time $t$. Thus, in the presence of take, the
population size on the right hand side of Eqn 1 is replaced by
$S(t) = X(t) - h(t)$.

To construct the value function, we consider a return when $X(t) = x(t)$
and harvest $h(t) = h$ denoted as the reward, $R(x(t), h)$. For example,
if the return is the harvest at time $t$, then
$R(x(t), h(t)) = \min(x(t), h(t))$. We assume that future harvests are
discounted relative to current ones at a constant rate of discount
$\delta$ and ask for the harvest policy that maximizes total discounted
harvest between the current time $t$ and a final time $T$. That is, we
seek to maximize over choices of harvest
$E [ \sum_{t = 0}^{T}  R(X(t), h(t), t) \delta^t]$, where the state
dynamics are given by Eqn 1 and $E$ denotes the expectation over future
population states.

@@

We consider a period of 40 in which data are obtained to estimate the
parameters or the GP. This is long enough that the estimates do not
depend on the particular realization, and longer times are not likely to
provide substantial improvement. Each of the models is fit to the same
data (Figure 1).

@@

The core difference for our purpose between the estimated GP and the
estimated parametric models is that the estimated GP model is defined
explicitly in reference to the observed data. As a result, uncertainty
arises in the GP model not only from uncertainty in the parameters, but
is also increases in regions farther from the observed states, such as
low population sizes in the example illustrated here. The estimated
parametric models, by contrast, are completely specified by the
parameters.

@@

A GP is a stochastic process for which any realization consisting of n
points follows a multivariate normal distribution of dimension $n$. To
characterize the GP we need a mean function and a covariance function.
We proceed as follows.

As before, we assume that the data $X(t)$ are observed with process
stochasticity around a mean function $g(X(t))$

@@

where $\varepsilon$ are IID normal random variables with zero-mean and
variance $\sigma^2$. Note that we have chosen to assume additive
stochasticity. While we could consider log-normal stochasticity as in
the parametric models, we make this choice to emphasize that the
Gaussian process approach need not have structurally correct
stochasticity to be effective.

In order to make predictions, we update the GP based on the observed set
of transitions. To do so, we collect the time series of observed states
into a vector of ``current'' states,
$\mathbf{X}_{\textrm{obs}} = \{X(1), \dots, X(T-1)\}$ and a vector of
``next'' states $\mathbf{Y}_{\textrm{obs}} = \{X(2),\dots,X(T)\}$ where
$T$ is the time of the final observation. Conditional on these
observations, the predicted next state, $g(X_p)$, for any given
``current'' state, $X_p$ follows a normal distribution with mean $E$ and
variance $C$ determined using the standard rules for conditioning in
multivariate normals, i.e.

\begin{equation}
E = K(\mathbf{X}_{\textrm p}, \mathbf{X}_{\textrm{obs}}) \left(K(\mathbf{X}_{\textrm{obs}},\mathbf{X}_{\textrm{obs}}) - \sigma \mathbb{I}_n \right)^{-1} \mathbf{Y}_{\textrm{obs}}
\end{equation}

and

\begin{equation}
C = K(\mathbf{X}_{\textrm p}, \mathbf{X}_{\textrm p}) - K(\mathbf{X}_{\textrm p}, \mathbf{X}_{\textrm{obs}}) \left(K(\mathbf{X}_{\textrm{obs}},\mathbf{X}_{\textrm{obs}}) - \sigma \mathbb{I} \right)^{-1} K(\mathbf{X}_{\textrm{obs}}, \mathbf{X}_{\textrm p})
\end{equation}

Here $\mathbb{I}_n is the $n\$ by $n$ identity matrix (i.e.~a matrix
with ones down the diagonal and zeros elsewhere) and $K$ is the
`covariance kernel.' The covariance kernel controls how much influence
one observation has on another. In the present application we use the
squared-exponential kernel which, when evaluated over a pair of vectors,
say $x$ and $y$, generates a covariance matrix whose $i,j$th element is
given by

\begin{equation}
K_{i,j}(\mathbf{x}, \mathbf{y}) = \exp\left( \frac{ -(x_i - y_j)^2}{2 \ell^2} \right)
\end{equation}

so that $\ell$ gives the characteristic length-scale over which
correlation between two observations decays. See Rasmussen and Williams
(2006) for other choices of covariance kernels and their properties.
Note that this simple formulation assumes a prior mean of zero. For the
parameters we use inverse Gamma priors on both the length-scale $\ell$
and $\sigma$, thus for example

@@

We derive the harvest policy from the estimated GP by inserting it into
a SDP algorithm. Given the GP posteriors, we construct the transition
matrix representing the probability of going to each state $X(t+1)$
given any current state $x(t)$ and any harvest $h(t)$ (See the function
\texttt{gp\_transition\_matrix()} in the provided R package). Given this
transition matrix, we use the same value iteration algorithm as in the
parametric case to determine the optimal policy. In doing so, the
uncertainty in the future state under the GP, $X(t+1)$, includes both
process uncertainty (based on the estimation of $\sigma$) and structural
uncertainty of the posterior collection of curves. Unlike the case of a
parametric model, which is a single curve given the parameters, the GP
remains a distribution of curves.

@@

To ensure our results are robust to the choice of parameters, we will
consider 96 different scenarios. To help better understand the process,
we first describe in detail the results of a single scenario.

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_1-1.pdf}
\caption{Points show the training data of stock-size over time. Curves
show the expectations over the posterior step-ahead predictions based on
each of the estimated models. Observe that all models are fitting the
data reasonably well.)}
\end{figure}

@@

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_2-1.jpeg}
\caption{The inferred Gaussian process compared to the true process and
maximum-likelihood estimated (MLE) process. We show the expected value
for the function $f$ under each model. Plots show replicates drawn from
the posterior distributions in order to convey uncertainty of the
estimates. Note the MLE is a point estimate of parameters and so
reflects no uncertainty in the distribution. The training data are also
shown as black points. The GP is conditioned on (0,0), shown as a
pseudo-data point.}
\end{figure}

We show the mean inferred population dynamics of each model relative to
the true model used to generate the data in Figure 2, predicting the
relationship between observed population size (x-axis) to the population
size after recruitment the following year.\\In addition to the raw data,
the GP is conditioned on going through the point 0,0 without error. All
parametric models also make this assumption. Conditioning on (0,0) is
equivalent to making the assumption that the population is closed, so
that once it hits 0 it stays at 0, despite the lack of any data in the
observed sequence to justify this. This assumption illustrates how the
GP can capture common-sense biology without having to assume more
explicit details about the dynamics at low population numbers that have
never been observed. If the population were not closed, one could repeat
the entire analysis without this assumption. Unlike parametric models,
the GP corresponds to a distribution of curves, of which this plot only
shows the means. Uncertainty in the parameters of the GP (not shown)
further widens the band of possible population sizes. In Figure S1 (see
supplement), we show the performance of the models outside the observed
training data.

@@

\begin{figure}[htbp]
\centering
\includegraphics{components/figure/manuscript-figure_4-1.jpeg}
\caption{In the management context, GPDP outperforms approaches based on
parametric models. We show 100 replicate simulations of the stock
dynamics (Eqn 1) under the policies derived from each of the estimated
models, as well as the policy based on the exact underlying model.}
\end{figure}

@@

Our results illustrate that model-choice approaches can be absolutely
misleading -- by providing support to models that cannot capture tipping
point dynamics because they have fewer parameters and the data are far
from the tipping point. That is, when the data come from around the
stable steady state, all the parametric models are approximately linear
and approximately identical. Thus, it is intuitive that all model
selection methods choose the simplest model. In a complex world, the
result is suboptimal. But in a world that might contain tipping points,
the result could be disastrous.

Many managers both in fisheries and beyond face a similar situation:
they have not observed the population dynamics at all possible
densities. A lack of comprehensive data at all population sizes,
combined with the inability to formulate accurate population models for
low population sizes in the absence of data, makes this situation the
rule more than the exception. Relying on parametric models and model
choice processes that favor simplicity ignores this basic reality. For a
long time, Carl Walters (e.g. Walters and Hilborn 1978) has argued that
if we began by fishing any newly exploited population down to very low
levels and then let it recover, we would be much better at estimating
population dynamics and thus predicting the optimal harvest levels.
While certainly true, this presents a rather risky policy in the face of
potential tipping points. The GPDP offers a risk-adverse alternative.

@@

Outside of the observed range of the data, the GP reverts to the prior,
and consequently the choice of the prior can also play a significant
role in determining the optimal policy. In the examples shown here we
have selected a prior that is both relatively uninformative (due to the
broad priors placed on its parameters $\ell$ and $\sigma$ and simple
(the choice of our covariance function, Eqns 12 and 13 ). In practice,
these should be chosen to confer particular biological properties. In
principle, this may allow a manager to improve the performance of the
GPDP by adding detail as is justified. For instance, it would be
possible to use a linear or a Ricker-shaped mean in the prior without
making the much stronger assumption that the Ricker is the structurally
correct model (Sugeno and Munch 2013a). One fruitful avenue of future
research is identifying criteria to ensure the prior and the reward
function are chosen appropriately for the problem at hand.

@@
